Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-08-22 review/response.Rmd

---
title: "Response"
author: "H. Sherry Zhang"
date: "22/08/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reviewer 1

## The Review

This work presents visual diagnostic tools for different methods of projection pursuit optimization. They use line plots, PCA and the tour method to visually compare the behavior of different optimizers. Finally, they showcased the analysis of three optimizers using their tool.

The tool is likely to be of interest to the R community.  Most of my comments are about the technical aspect of the content. The writing is overall good, but please see my comment under the "writing" regarding to the clarity of the text.

### Major comment

- "Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign...However, this creates difficulties for comparing the optimisers since the end bases will be symmetric to the origin. A sign flipping step is conducted..." and the "Reconciling the orientation" section
  - For 1D projections, flipping the sign is sufficient. For p-D projections of $\mathbb{R}^n$​, one may need a little more. To my knowledge most index functions are invariant to any rotations that happens within the p-subspace. This makes me wonder that in general, one may want to align two bases so that their difference does not contain any "within-subspace" components. In other words, one may want to compare representations in the Grassmannian manifold  (space of k-subspaces) instead of in the Stiefel manifold (space of orthonormal k-frames).

\textcolor{red}{Is this a comment or anything to change?}

### Minor comment

- The software seems to be specific to projection pursuit. I wish the authors could comment more on its potential generality to other optimizations. For example, can one generalizes the tour method to visualize any parameter trace of an optimizer?

**It is feasible to generalise the tour method for diagnosing other optimisers, providing the parameter space can be simulated. In our examples, the parameter to trace (projection base ($p \times d$) ) lives in an p-sphere, which can be simulated by the `geozoo` package. **

- Figure 4 "...All the bases in PD have been flipped for easier comparison of the final bases..."
  - This is related to the major comment. In the figure, PD flipped in sign but stopped at a basis representation similar to the destination of the CRS. Flipping the sign back, does it mean that PD and CRS diverged but stopped at the similar subspace with opposite basis representations? Also, given that PD and CRS have the same starting bases and one is flipped, why wouldn't the dashed line pass the origin (gray dot) in the figure?

\textcolor{red}{The grey dot is the the zero matrix after projecting by the PCA. }

- "Recall that the columns in a 2D basis are orthogonal to each other, so the space of *p* × 2 bases is a *p*-D torus"
  - It is not clear to me why "so". For p=2, the space of 2x2 bases is a pair of 1-D tori instead of just one (because you have two orientations for a pair of orthogonal vectors), right? For p=3, the space of 3x2 bases is SO(3), which is not a 3-D torus? Ref: https://en.wikipedia.org/wiki/Stiefel_manifold

\textcolor{red}{TODO}

### Writing

- "Along with the bases recorded during the optimisation and a zero basis, the parameters (centre and radius) of the 2D space can be estimated by PCA. The centre of the circle is the first two PC coordinates of the zero matrix, and the radius is estimated as the largest distance from the centre to the basis."
  - It took me multiple passes to get what this means. Perhaps instead starting with something like: "We use PCA to project and visualize the parameters/bases in 2D. The centre of the 2D view is..."?

**Changed to "We use PCA to estimate the 2D space that contains all the bases recorded and the zero basis. The centre of the 2D view is the first two PC coordinates of the zero matrix, and the radius is estimated as the largest distance from the centre to the basis."**

- "Figure 6 shows some frames from the tour plot of the same two optimisations in its original 5D space."
  - I think this is the first mention of "5D space"/5D data in the main text (apart from the caption in Figure 4), so it may be worth clarifying what the original 5D space is.

\textcolor{red}{The description of data is in the subsection simulation setup in the next section.}

- "While the previous few examples have looked at the space of 1D bases ...  2D bases"
  - I don't think the terms "1D bases" and "2D bases" are standard. Perhaps (orthonormal) "1-frames" and "2-frames"?

**Changed to "While the previous few examples have looked at the space of 1-frames in a unit sphere, this section visualises the space of 2-frames."**

# Reviewer 2

These experiments are interesting and so is the methodology.
The conclusions could do with beefing up
What can we learn about how to optimise the holes index for 1 and 2d?
Does this apply to other smooth indices? What about non-smooth indices?
What about indices in other dimensions?

\textcolor{red}{Is the reviewer suggesting adding more details in the first paragraph of the conclusion?}

1. Should refer to earlier work on scagnostic indices, page 3

\textcolor{red}{my understanding is that although scagnostics were introduced much earlier, using it as an index function for tourr is a recent implemntation.}

2. figure 2 caption "using the holes index"  "and a boxplot". What is the colour scale for the points? The text related to figure 2 in the bottom of page 5 should mention the dataset and the index, giving references if needed.

**Fixed language. Added mention the color scale in the figure 2 caption:  "The color scale is from the customised botanical palette in the ferrn package."**

3. "with higher index value"  -> "with a higher index value". The word "the" is missing in a few other places in the paper.

**corrected.**

4. In Figure 3 would be better if the y-axes scales were the same for the two graphs.

**fixed**

5. What is the star in figures 4 and 5? This is not mentioned in the text or caption

**Added "The black star is the theoretical best basis the optimisers aiming to find." in the caption of figure 4.**

6. In the text you say things like "details=TRUE" and "animate=TRUE" but what functions are you referring to?

**Mention of "details=TRUE" and "animate=TRUE" is removed from subsection diagnostic 3a and 3b to the implementation section: `explore_space_pca()` produces the PCA plot of projection bases on the reduced space. Figure 4 includes the additional details of anchor and search bases, which can be turned on by the argument `details = TRUE`. The animated version in Figure 5 is produced with argument `animate = TRUE`.**

7. In Figure 9, it seems CRS is doing better here than in Figure 8-- maybe this could be mentioned.

**Figure 8 and 9 focuses on different aspects: Figure 8 compares the effect of implementing an interruption while Figure 9 compares the effect of polishing. Cross comparing the two experiments is less meaningful.**

8. "not the case for all the indexes" missing a "." page 11

**added**

9. For the Kolmogorov index, why are you using randomly generated data instead of normal quantiles?

\textcolor{red}{TODO}

10. Fix references with "et al"

**fixed**

### 2021-08-22 paper/zhang-cook-laa-langrene-menendez.Rmd
changes made as per the reviewers

```{r toy-search, fig.cap="A comparison of the searches by two optimisers: CRS (left) and SA (right) on a 2D projection problem of a six-variable dataset, \\code{boa6} using holes index. Both optimisers reach the final basis with a similar index value while it takes SA longer to find the final basis. In the earlier iterations, optimisers quickly find a better basis to proceed while in the later iterations, most sampled bases fail to make an improvement on the index value and boxplot is used to summarise the distribution of the index values. There is no better basis found in the last iteration, 9 (left) and 15 (right), before reaching the maximum number of try and hence it is colored grey."}
```

@@

```{r toy-interp, fig.cap = "An inspection of the index values as the optimisation progress for two optimisers: CRS (left) and SA (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \\code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SA is observed to first pass by some bases with higher index value before reaching the target bases in time 76-130."}
```

Another interesting feature to examine is the changes in the index value between interpolating bases since the projection on these bases is shown in the tour animation. Trace plots are created by plotting the index value against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisations as Figure \ref{fig:toy-search} and one can observe that the trace is smooth in both cases. It may seem bizarre at first sight that the interpolation sometimes passes bases with higher index value before it decreases to a lower target. This happens because, on the one hand, the probabilistic acceptance in SA implies that some worse bases will be accepted by the optimiser. In addition, the guided tour interpolates between the current and target basis to provide a smooth transition between projections, and sometimes a higher index value will be observed along the interpolation path. This indicates that a non-monotonic interpolation cannot be avoided, even for CRS. Later, in section \ref{monotonic}, there will be a discussion on improving the non-monotonic interpolation for CRS.

@@

```{r toy-pca, fig.height = 4, fig.asp = 1, fig.cap = "Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD have been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

Apart from checking the search and progression of an optimiser, looking at where the bases are positioned in the basis space is also of interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a Stiefel manifold. For one-dimensional projections, this forms a $p$-dimensional sphere. A dimensionality reduction method, e.g. principal component analysis is applied to first project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various types of bases involved: 1) The starting basis; 2) The search bases that the optimiser evaluated to produce the anchor bases; 3) The anchor bases that have the highest index value in each iteration; 4) The interpolating bases on the interpolation path; and finally, 5) the end basis. The importance of these bases differs and the most important ones are the starting, interpolating, and end bases. The anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the orientation of the basis and so is the index value. However, this creates difficulties for comparing the optimisers since the end bases will be symmetric to the origin. A sign flipping step is conducted to flip the signs of all the bases in one routine if different optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As mentioned previously, the original basis space is a high-dimensional sphere and random bases on the sphere can be generated via the \CRANpkg{geozoo} [@geozoo] package. Along with the bases recorded during the optimisation and a zero basis, the parameters (centre and radius) of the 2D space can be estimated by PCA. The centre of the circle is the first two PC coordinates of the zero matrix, and the radius is estimated as the largest distance from the centre to the basis. The theoretical best basis is known for simulated data and can be labelled to compare how close the end basis found by the optimisers is to the theoretical best. Various aesthetics, i.e. size, alpha (transparency), and colour, are applicable to emphasize critical elements and adjust for the presentation. For example, anchor points and search points are less important and hence a smaller size and alpha are used. Alpha can also be applied on the interpolation paths to show the start to finish from transparent to opaque.

@@

An animation is another type of display to show how the search progresses from start to finish in the space. An  `animate = TRUE` argument is used to enable the animation in the PCA plot. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that CRS finds its end basis quicker than PD since CRS finishes its search in the 5th frame while PD is still making more progress.

@@

While the previous few examples have looked at the space of 1D bases in a unit sphere, this section visualises the space of 2D bases. Recall that the columns in a 2D basis are orthogonal to each other, so the space of $p \times 2$ bases is a $p$-D torus. For $p = 3$ one would see a classical 3D torus shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two circles of the torus can be observed to be perpendicular to each other and this can be linked back to the orthogonality condition. Two paths from CRS and PD are plotted on top of the torus and coloured in green and brown respectively to match the previous plots. The final basis found by PD and CRS are shown in a larger shape  and printed below respectively:

@@

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. For SA, a non-monotonic interpolation is justified since target bases do not necessarily have a higher index value than the current one, while this is not the case for CRS. The original trace plot for a 2D projection problem, optimised by CRS, is shown on the left panel of Figure \ref{fig:interruption} and one can observe that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop at the best basis found in the interpolation. The right panel of Figure \ref{fig:interruption} shows the trace plot after implementing the interruption and while the first two interpolations are identical, the basis at time 61 has a higher index value than the target in the third interpolation. Rather than starting the next iteration from the target basis on time 65, CRS starts the next iteration at time 61 on the right panel and reaches a better final basis.

```{r interruption, fig.cap = "Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \\code{boa6} data using holes index, optimised by CRS. On the left panel, basis with higher index value is found during the interpolation but not used. On the right panel, the interruption stops the interpolation at the basis with the highest index value for each iteration and results in a final basis with higher index value, as shown on the right panel."}
```

@@

The holes index function used for all the examples before this section produces a smooth interpolation, while this is not the case for all the indexes An example of a noisy index function for 1D projections compares the projected data, $\mathbf{Y}_{n \times 1}$, to a randomly generated normal distribution, $\mathcal{N}_{n \times 1}$, using the Kolmogorov test. Let $F_{.}(n)$ be the ECDF function (empirical cummulative distribution function), with two possible subscripts $Y$ and $\mathcal{N}$ representing the projected and randomly generated data, and $n$ denoting the number of observation, the Normal Kolmogorov index $I^{nk}(n)$, is defined as:

@@

  - \code{explore\_trace\_search()} produces summary plots in Figure \ref{fig:toy-search}.
  - \code{explore\_trace\_interp()} produces trace plots for the interpolation points in Figure \ref{fig:toy-interp}.
  - \code{explore\_space\_pca()} produces the PCA plot of projection bases on the reduced space in Figure \ref{fig:toy-pca}. The animated version in Figure \ref{fig:toy-pca-animated} can be turned on via the argument `animate = TRUE`.
  - \code{explore\_space\_tour()} produces animated tour view on the full space of the projection bases in Figure \ref{fig:toy-tour}.

>>>>

```{r toy-search, fig.cap="A comparison of the searches by two optimisers: CRS (left) and SA (right) on a 2D projection problem of a six-variable dataset, \\code{boa6} using the holes index. Both optimisers reach the final basis with a similar index value while it takes SA longer to find the final basis. In the earlier iterations, optimisers quickly find a better basis to proceed while in the later iterations, most sampled bases fail to make an improvement on the index value and a boxplot is used to summarise the distribution of the index values. There is no better basis found in the last iteration, 9 (left) and 15 (right), before reaching the maximum number of try and hence it is colored grey. The color scale is from the customised botanical palette in the ferrn package."}
```

@@

```{r toy-interp, fig.cap = "An inspection of the index values as the optimisation progress for two optimisers: CRS (left) and SA (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \\code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SA is observed to first pass by some bases with higher index values before reaching the target bases in time 76-130."}
```

Another interesting feature to examine is the changes in the index value between interpolating bases since the projection on these bases is shown in the tour animation. Trace plots are created by plotting the index value against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisations as Figure \ref{fig:toy-search} and one can observe that the trace is smooth in both cases. It may seem bizarre at first sight that the interpolation sometimes passes bases with higher index values before it decreases to a lower target. This happens because, on the one hand, the probabilistic acceptance in SA implies that some worse bases will be accepted by the optimiser. In addition, the guided tour interpolates between the current and target basis to provide a smooth transition between projections, and sometimes a higher index value will be observed along the interpolation path. This indicates that a non-monotonic interpolation cannot be avoided, even for CRS. Later, in section \ref{monotonic}, there will be a discussion on improving the non-monotonic interpolation for CRS.

@@

```{r toy-pca, fig.height = 4, fig.asp = 1, fig.cap = "Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. The black star is the theoretical best basis the optimisers aiming to find. All the bases in PD have been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

Apart from checking the search and progression of an optimiser, looking at where the bases are positioned in the basis space is also of interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a Stiefel manifold. For one-dimensional projections, this forms a $p$-dimensional sphere. A dimensionality reduction method, e.g. principal component analysis is applied to first project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various types of bases involved: 1) The starting basis; 2) The search bases that the optimiser evaluated to produce the anchor bases; 3) The anchor bases that have the highest index value in each iteration; 4) The interpolating bases on the interpolation path; and finally, 5) the end basis. The importance of these bases differs and the most important ones are the starting, interpolating, and end bases. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the orientation of the basis and so is the index value. However, this creates difficulties for comparing the optimisers since the end bases will be symmetric to the origin. A sign flipping step is conducted to flip the signs of all the bases in one routine if different optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As mentioned previously, the original basis space is a high-dimensional sphere and random bases on the sphere can be generated via the \CRANpkg{geozoo} [@geozoo] package. We use PCA to estimate the 2D space that contains all the bases recorded and the zero basis. The centre of the 2D view is the first two PC coordinates of the zero matrix, and the radius is estimated as the largest distance from the centre to the basis. The theoretical best basis is known for simulated data and can be labelled to compare how close the end basis found by the optimisers is to the theoretical best. Various aesthetics, i.e. size, alpha (transparency), and colour, are applicable to emphasize critical elements and adjust for the presentation. For example, anchor points and search points are less important and hence a smaller size and alpha are used. Alpha can also be applied on the interpolation paths to show the start to finish from transparent to opaque.

@@

An animation is another type of display to show how the search progresses from start to finish in the space. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that CRS finds its end basis quicker than PD since CRS finishes its search in the 5th frame while PD is still making more progress.

@@

While the previous few examples have looked at the space of 1-frames in a unit sphere, this section visualises the space of 2-frames. Recall that the columns in a 2D basis are orthogonal to each other, so the space of $p \times 2$ bases is a $p$-D torus. For $p = 3$ one would see a classical 3D torus shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two circles of the torus can be observed to be perpendicular to each other and this can be linked back to the orthogonality condition. Two paths from CRS and PD are plotted on top of the torus and coloured in green and brown respectively to match the previous plots. The final basis found by PD and CRS are shown in a larger shape  and printed below respectively:

@@

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with a higher index value than the target one. For SA, a non-monotonic interpolation is justified since target bases do not necessarily have a higher index value than the current one, while this is not the case for CRS. The original trace plot for a 2D projection problem, optimised by CRS, is shown on the left panel of Figure \ref{fig:interruption} and one can observe that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop at the best basis found in the interpolation. The right panel of Figure \ref{fig:interruption} shows the trace plot after implementing the interruption and while the first two interpolations are identical, the basis at time 61 has a higher index value than the target in the third interpolation. Rather than starting the next iteration from the target basis on time 65, CRS starts the next iteration at time 61 on the right panel and reaches a better final basis.

```{r interruption, fig.cap = "Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \\code{boa6} data using holes index, optimised by CRS. On the left panel, basis with a higher index value is found during the interpolation but not used. On the right panel, the interruption stops the interpolation at the basis with the highest index value for each iteration and results in a final basis with a higher index value, as shown on the right panel."}
```

@@

The holes index function used for all the examples before this section produces a smooth interpolation, while this is not the case for all the indexes. An example of a noisy index function for 1D projections compares the projected data, $\mathbf{Y}_{n \times 1}$, to a randomly generated normal distribution, $\mathcal{N}_{n \times 1}$, using the Kolmogorov test. Let $F_{.}(n)$ be the ECDF function (empirical cummulative distribution function), with two possible subscripts $Y$ and $\mathcal{N}$ representing the projected and randomly generated data, and $n$ denoting the number of observation, the Normal Kolmogorov index $I^{nk}(n)$, is defined as:

@@

  - \code{explore\_trace\_search()} produces summary plots in Figure \ref{fig:toy-search}.
  - \code{explore\_trace\_interp()} produces trace plots for the interpolation points in Figure \ref{fig:toy-interp}.
  - \code{explore\_space\_pca()} produces the PCA plot of projection bases on the reduced space in Figure \ref{fig:toy-pca}. The animated version in Figure \ref{fig:toy-pca-animated} can be turned on via the argument `animate = TRUE`.
   - \code{explore\_space\_pca()} produces the PCA plot of projection bases on the reduced space. Figure \ref{fig:toy-pca} includes the additional details of anchor and search bases, which can be turned on by the argument `details = TRUE`. The animated version in Figure \ref{fig:toy-pca-animated} is produced with argument `animate = TRUE`.
  - \code{explore\_space\_tour()} produces animated tour view on the full space of the projection bases in Figure \ref{fig:toy-tour}.
